Learning to Pivot with Adversarial Networks

### 2017-05-15 tex/nips/nips_2017.tex
save

\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{graphicx}
\usepackage{subfigure}
\usepackage{natbib}
\usepackage{algorithm}
\usepackage{algorithmic}

@@

\begin{abstract}
    Many inference problems involve data generation processes that are not uniquely
    specified or are uncertain in some way. In a scientific context, the presence of
    several plausible data generation processes is often associated to the presence
    of systematic uncertainties. Robust inference is possible if it is based on a
    pivot -- a quantity whose distribution does not depend on the unknown value of the
    (categorical or continuous) nuisance parameters that parametrizes this family of
    generation processes. In this work,  we introduce a flexible training procedure
    based on adversarial networks for enforcing the pivotal property on a predictive model, i.e., its independence
    with respect to nuisance parameters. We derive theoretical results showing that the proposed algorithm tends
    towards a minimax solution corresponding to a predictive model that is both
    optimal and independent of the nuisance parameters (if that model exists) or
    for which one can tune a trade-off between power and robustness. Finally, we
    demonstrate the effectiveness of this approach with a toy example and an example
    from particle physics.
\end{abstract}

@@

In this work, we focus on supervised learning techniques where the presence of
systematic uncertainty can be associated to the fact that the data generation
process is not uniquely specified. In other words, the lack of systematic
uncertainties corresponds to the (rare) case that the process that generates
training data is unique, fully specified, and an accurate representative of the
real world data. By contrast, a common situation when systematic uncertainty is
present is when the training data are not representative of the real data that a
predictive model will be applied to in practice.
Several techniques for domain
adaptation have been developed to create predictive models that are more robust
to this type of uncertainty.
A more generic situation is that there are several
plausible data generation processes, specified as a family of data generation
processes parametrized by nuisance parameters, which can be categorical or
continuous.
In this broader context, statisticians have for long been working on the development of robust inference techniques
based on the concept of a pivot -- a quantity whose distribution is invariant
with the nuisance parameters. Pivotal quantities are e.g. fundamental to the construction
of test statistics and confidence intervals.

@@

The remainder of the paper is structured as follows. In Sec.~\ref{sec:problem},
we first formally define the problem of learning a pivotal, predictive model. In
Sec. \ref{sec:method} and \ref{sec:theory}, we describe how adversarial
networks can be trained to obtain predictive models that are independent of
nuisance parameters and show that the proposed procedure converges
towards a model that is both optimal and robust. The  effectiveness of the
approach is then demonstrated empirically in Sec. \ref{sec:toy} and \ref{sec:hep}.
In Sec.~\ref{sec:related}, we relate our work to
close contributions from statistical inference, domain adaptation, and fairness
in classification. Finally, we gather our concluding remarks in
Sec.~\ref{sec:conclusions}.

@@

We begin with a family of data generation processes $p(X,Y,Z)$, where $X \in
\mathcal{X}$ are the data, $Y\in \mathcal{Y}$ are the target labels, and $Z\in
\mathcal{Z}$ are the nuisance parameters that can be continuous or categorical.
% We assume training data $\{x_i, y_i, z_i\}_{i=1}^N$.
Let us assume that prior to incorporating the effect of uncertainty
in $Z$, our goal is to learn a regression function $f : \mathcal{X} \mapsto
{\cal S}$ of parameters $\theta_f$ (e.g., a neural network-based probabilistic
classifier) that minimizes a loss ${\cal L}_f(\theta_f)$ (e.g., the
cross-entropy).

We augment our initial objective so that inference based on $f(X ; \theta_f)$ will be
robust to the value $z \in {\cal Z}$ of the nuisance parameter $Z$  -- which remains unknown at
test time. A formal way of enforcing robustness is to require that the distribution of
$f(X ; \theta_f)$ conditional on $Z$ (and possibly $Y$) be invariant with
 the nuisance parameter $Z$. Thus, we wish to find a function $f$ such that
\begin{equation}\label{eqn:criterion}
    p(f(X ; \theta_f) = s | z ) = p(f(X ; \theta_f) = s | z^\prime )
\end{equation}
for all $z,z^\prime \in  {\cal Z}$ and all values $s \in {\cal S}$ of $f(X ; \theta_f)$.
In words, we are looking for a predictive function $f$
which is a pivotal quantity (see e.g., \citep{degroot1986probability}) with respect to the
nuisance parameters. This implies that  $f(X; \theta_f)$ and $Z$ are independent random variables.

@@

If $Z$ takes discrete values, then $p_{\theta_r}$ can be represented e.g. as a
probabilistic classifier $\mathbb{R} \mapsto \mathbb{R}^{|{\cal Z|}}$ whose
$j^\textrm{th}$ output (for $j=1, \dots, |{\cal Z}|$) is the estimated probability mass
$p_{\theta_r}(z_j|f(X;\theta_f)=s)$. Similarly, if $Z$ takes continuous values,
then we can model the posterior probability density $p(z | f(X;\theta_f)=s)$
with a sufficiently flexible  parametric family of distributions $\mathcal{P}(\gamma_1, \gamma_2, \dots)$,
where the parameters $\gamma_j$ depend on $f(X, \theta_f)$ and $\theta_r$.
The adversary $r$ may
take any form, i.e. it does not need to be a neural network, as long as it exposes a
differentiable function $p_{\theta_r}(z|f(X;\theta_f)=s)$ of sufficient capacity
to represent the true distribution. Fig.~\ref{fig:architecture} illustrates
a concrete example where $p_{\theta_r}(z | f(X;\theta_f)=s)$ is a mixture of
gaussians, as modeled with a mixture density network~\citep{bishop1994mixture}).
The $j^\textrm{th}$ output corresponds to the estimated value of the
corresponding parameter $\gamma_j$ of that distribution (e.g., the mean,
variance and mixing coefficients of its components). As in
\citep{nix1994estimating,bishop1994mixture}, the estimated probability density
$p_{\theta_r}(z|f(X;\theta_f)=s)$ can then be evaluated for any $z \in {\cal Z}$ and any score $s \in {\cal S}$.

>>>>

\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{graphicx}
\usepackage{subfigure}
\usepackage{algorithm}
\usepackage{algorithmic}

@@

\begin{abstract}
    Many inference problems involve data generation processes that are not uniquely
    specified or are uncertain in some way. In a scientific context, the presence of
    several plausible data generation processes is often associated to the presence
    of systematic uncertainties. Robust inference is possible if it is based on a
    pivot -- a quantity whose distribution does not depend on the unknown value of the
    (categorical or continuous) nuisance parameters that parametrizes this family of
    generation processes. In this work,  we introduce a training procedure
    based on adversarial networks for enforcing the pivotal property on a predictive model, i.e., its independence
    with respect to nuisance parameters. We derive theoretical results showing that the proposed algorithm tends
    towards a minimax solution corresponding to a predictive model that is both
    optimal and independent of the nuisance parameters (if that model exists) or
    for which one can tune a trade-off between power and robustness. Finally, we
    demonstrate the effectiveness of this approach with a toy example and an example
    from particle physics.
\end{abstract}

@@

In this work, we focus on supervised learning techniques where the presence of
systematic uncertainty can be associated to the fact that the data generation
process is not uniquely specified. In other words, the lack of systematic
uncertainties corresponds to the (rare) case that the process that generates
training data is unique, fully specified, and an accurate representative of the
real world data. By contrast, a common situation when systematic uncertainty is
present is when the training data are not representative of the real data that a
predictive model will be applied to in practice.
Several techniques for domain
adaptation have been developed to create predictive models that are more robust
to this type of uncertainty.
A more generic situation is that there are several
plausible data generation processes, specified as a family of data generation
processes parametrized by nuisance parameters, which can be categorical or
continuous.
In this broader context, statisticians have for long been working on the development of robust inference techniques
based on the concept of a pivot -- a quantity whose distribution is invariant
with the nuisance parameters (see e.g., \citep{degroot1986probability}). Pivotal quantities are e.g. fundamental to the construction
of test statistics and confidence intervals.

@@

% % this adds nothing
% The remainder of the paper is structured as follows. In Sec.~\ref{sec:problem},
% we first formally define the problem of learning a pivotal, predictive model. In
% Sec. \ref{sec:method} and \ref{sec:theory}, we describe how adversarial
% networks can be trained to obtain predictive models that are independent of
% nuisance parameters and show that the proposed procedure converges
% towards a model that is both optimal and robust. The  effectiveness of the
% approach is then demonstrated empirically in Sec. \ref{sec:toy} and \ref{sec:hep}.
% In Sec.~\ref{sec:related}, we relate our work to
% close contributions from statistical inference, domain adaptation, and fairness
% in classification. Finally, we gather our concluding remarks in
% Sec.~\ref{sec:conclusions}.

@@

We begin with a family of data generation processes $p(X,Y,Z)$, where $X \in
\mathcal{X}$ are the data, $Y\in \mathcal{Y}$ are the target labels, and $Z\in
\mathcal{Z}$ are the nuisance parameters that can be continuous or categorical.
% We assume training data $\{x_i, y_i, z_i\}_{i=1}^N$.
Let us assume that prior to incorporating the effect of uncertainty
in $Z$, our goal is to learn a regression function $f : \mathcal{X} \mapsto
{\cal S}$ of parameters $\theta_f$ (e.g., a neural network-based probabilistic
classifier) that minimizes a loss ${\cal L}_f(\theta_f)$ (e.g., the
cross-entropy). In classification, values $\smash{s \in {\cal S} = \mathbb{R}^{|{\cal Y}|}}$
correspond to scores used for mapping hard predictions $y \in {\cal Y}$, while ${\cal S} = {\cal Y}$ for regression.

We augment our initial objective so that inference based on $f(X ; \theta_f)$ will be
robust to the value $z \in {\cal Z}$ of the nuisance parameter $Z$  -- which remains unknown at
test time. A formal way of enforcing robustness is to require that the distribution of
$f(X ; \theta_f)$ conditional on $Z$ (and possibly $Y$) be invariant with
 the nuisance parameter $Z$. Thus, we wish to find a function $f$ such that
\begin{equation}\label{eqn:criterion}
    p(f(X ; \theta_f) = s | z ) = p(f(X ; \theta_f) = s | z^\prime )
\end{equation}
for all $z,z^\prime \in  {\cal Z}$ and all values $s \in {\cal S}$ of $f(X ; \theta_f)$.
In words, we are looking for a predictive function $f$
which is a pivotal quantity with respect to the
nuisance parameters. This implies that  $f(X; \theta_f)$ and $Z$ are independent random variables.

@@

If $Z$ takes discrete values, then $p_{\theta_r}$ can be represented as a
probabilistic classifier $\mathbb{R} \mapsto \mathbb{R}^{|{\cal Z|}}$ whose
$j^\textrm{th}$ output (for $j=1, \dots, |{\cal Z}|$) is the estimated probability mass
$p_{\theta_r}(z_j|f(X;\theta_f)=s)$. Similarly, if $Z$ takes continuous values,
then we can model the posterior probability density $p(z | f(X;\theta_f)=s)$
with a sufficiently flexible  parametric family of distributions $\mathcal{P}(\gamma_1, \gamma_2, \dots)$,
where the parameters $\gamma_j$ depend on $f(X, \theta_f)$ and $\theta_r$.
The adversary $r$ may
take any form, i.e. it does not need to be a neural network, as long as it exposes a
differentiable function $p_{\theta_r}(z|f(X;\theta_f)=s)$ of sufficient capacity
to represent the true distribution. Fig.~\ref{fig:architecture} illustrates
a concrete example where $p_{\theta_r}(z | f(X;\theta_f)=s)$ is a mixture of
gaussians, as modeled with a mixture density network~\citep{bishop1994mixture}).
The $j^\textrm{th}$ output corresponds to the estimated value of the
corresponding parameter $\gamma_j$ of that distribution (e.g., the mean,
variance and mixing coefficients of its components). The estimated probability density
$p_{\theta_r}(z|f(X;\theta_f)=s)$ can then be evaluated for any $z \in {\cal Z}$ and any score $s \in {\cal S}$.
